Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-14 paper/paper.tex
Current version

This paper presents the first Coq formalisation of
a recursion scheme, called the \emph{hylomorphism}, along with its algebraic
laws allowing for the mechanisation of all recognised (terminating) recursive
algorithms. The key contribution of this paper is that this formalisation is
fully axiom-free allowing for the extraction of safe, idiomatic OCaml code. We
exemplify the framework by formalising a series of algorithms based on different
recursive paradigms such as divide-and conquer, dynamic programming, and mutual
recursion and demonstrate that the extracted OCaml code for the programs
formalised in our framework is efficient, resembles code that a human programmer
would write, and contains no occurrences of \ocaml{Obj.magic}.  We also present
a machine-checked proof of the well-known short-cut fusion optimisation.
\end{abstract}

@@

Despite the rigorous set of algebraic laws that are satisfied by recursion
schemes, there is a lack of tool support for their use in the context of program
calculation. In fact, most of the work on applying program calculation is done
by performing pen-and-paper proofs, and then translating the result to specific
instances of recursion schemes, implemented in a programming language (generally
Haskell). There are some examples of implementations of program calculation
techniques, but these implementations are scarce, not up to date, and not
verified in a proof assistant. For example, Cunha et al.~\cite{DrHylo} automated
program calculation techniques by a custom Haskell implementation.

Several authors build tools for applying program calculation techniques by
\emph{mechanising} them as part of a proof assistant. However, most of the work
in mechanising recursion schemes focuses on a narrow subset of the known
recursion schemes due to termination issues, or do not focus on proving
algebraic laws of generic recursion
schemes~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.  Indeed,
most of the mechanisations of generic recursion schemes focus on \emph{maps} and
\emph{folds}, and few authors focus on the \emph{unfolds}. Omitting
\emph{unfolds} severely limits the expressivity of the resulting mechanisations,
and prevent them from being able to mechanise the most general recursion scheme:
a divide-and-conquer algorithm which goes under the name of
\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}. This generality has been proven by
Hinze et al.~\cite{HinzeWG15}, by showing that \emph{all known  recursion
schemes are instances of hylomorphisms}.  To date, no work has mechanised
hylomorphisms in the Coq proof assistant, together with their algebraic laws.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code, and it is not well-suited for
program calculation. This is unfortunate since code extration is what allows the
execution of code that has been verified in
Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
al's approach, extraction (1) does not preserve the recursive structure of
common implementations; and (2) leads to unsafe casts like \ocaml{Obj.magic} in
the generated code. This latter is also problematic in that, for higher-order
programs, simple interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

The contributions of this paper are as follows.  This work presents the first
Coq formalisation \emph{hylomorphisms} that (1) is \emph{fully axiom-free}; (2)
allows the extraction of idiomatic OCaml code; and (3) can use regular Coq
equalities to do program calculation, derive correct implementations, and apply
optimisations.
The full mechanisation is open source.
%can be found on
%Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
%  ANONYMISE THIS}.
While programmers still need to reason about the termination of their programs,
the use \emph{program calculation} can ease this task by allowing the use of
fusion on programs that have already been proven to terminate.
% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}
%

@@

\section{Recursion Schemes}\label{sec:rec-schemes}
The structure of data is very similar to the structure of an
algorithm which processes that data. This relationship manifests in the form of
structured recursion schemes which are widely used in functional languages such
Haskell. Canonical examples are folds (catamorphisms), which consume data, and
unfolds (anamorphisms), which produce it. While some implementations of
recursion schemes like $\operator{foldr}$ in Haskell are specific to a particular
data structure, in this case $\operator{List}$s, we can generalise further these
ideas to account for generic (co)inductive data types and generic algorithms
operating on them.

@@

In $\Set$, we can define the set of lists
\[
  \List{A} \iso 1 + A \times \List{A}
\]
as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$.
The terminal object in $\Set$ is the \emph{unit} type, that only has one
inhabitant, $* : 1$.
The $\List{-}$ type is a functor. Its action on objects is to take any set
$A$ to $\List{A}$, and its functorial
action $\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument.
However, there is a more general way of proving these type of definitions are
well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
the greatest set generated by the constructor
$\cons : A \times \Str{A} \to \Str{A}$.

\subsubsection{Functors in Haskell}
It is often helpful for functional programmers to think of categorical
constructions in terms of Haskell thereby assuming that Haskell types are sets
and programs are functions~\cite{DanielssonHJG06}.

We can declare the class of types which are functors. This is done by imposing
that a function \mintinline{haskell}{f : * -> *} comes with a functorial action
given by function \mintinline{haskell}{fmap}:
\begin{minted}{haskell}
class Functor f where
  fmap :: (a -> b) -> f a -> f b
\end{minted}

We now define the type of lists as usual and then show that it is an instance of
the class \mintinline{haskell}{Functor} by implementing the
\mintinline{haskell}{fmap} program:
\begin{minted}{haskell}
data List a = Nil | Cons a (List a)

instance Functor List where
 fmap f Nil = Nil
 fmap f (Cons x xs) = Cons (f x) (fmap f xs)
\end{minted}

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
where $X$ is an object of the category called the \emph{carrier} of the algebra
and $a_{X}$ is an arrow of type $FX \to X$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
\emph{$F$-algebra homomorphisms}  $f : X \to Y$ in $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ a_{X} = A_{Y} \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
the initial $F$-algebra is precisely the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ with
$\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
isomorphism $F\mu F \iso \mu F$.  By the property of initial $F$-algebras, for
any other $F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra
homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (X, a_{X})$ and
pronounced ``\emph{catamorphism}'' satisfying:
\begin{equation}
  \label{eq:cata:unique}
  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InIso
\end{equation}
In words, any $F$-algebra homomorphism from the initial $F$-algebra is a
catamorphism. As a result of the uniqueness property we can derive the fusion
law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
\begin{equation}
  \label{eq:cata:fusion}
  f \circ \cata{a_{X}} = \cata{a_{Y}}
\end{equation}
which means that the composition of a program $f$ with a catamorphism recursing
once over the data structure is the same as performing that recursion once using
the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
optimisation as we shall see.

@@

\subsubsection{Inductive Types and Catamorphisms in Haskell }
We now implment the least fixed-point of a functor \mintinline{haskell}{f}.
Inuitively this is the type of \mintinline{haskell}{f}-branching trees of finite depth:
\begin{minted}{haskell}
newtype Fix f = In {inOp :: f (Fix f )}
\end{minted}
Note that since Haskell is a lazy language with general recursion this type can
also represent infinite trees. However, the point of using recursion schemes is
precisely that of avoid using general recursion in the first place.

A catamorphism can be readily implemented as follows
\begin{minted}{haskell}
cata :: Functor f => (f a -> a) -> Fix f -> a
cata alg = alg . fmap (cata alg) . In
\end{minted}

As an example of how to construct inductive types we define the functor
$X \mapsto 1 + A \times X$ as data type parameterized by a type:
\begin{minted}{haskell}
data ListF a x = Nil | Cons a deriving Functor
newtype List a = (Fix ListF) a
\end{minted}
It should be easy now to see that a foldr is precisely a catamorphism over the
type \mintinline{haskell}{List} which we have just defined.  For example,
summing up the elements of a list can be implemented using a catamorphism as
follows
\begin{minted}{haskell}
sum :: Fix (ListF Int) -> Int
sum xs = cata (\case { Nil -> 0; Cons n m -> n + m } )  xs
\end{minted}
In the program above, the code inside the lambda case is precisely the
\mintinline{haskell}{ListF Int}-algebra for the catamorphism.

@@

\subsubsection{Coinductive Types and Anamorphisms in Haskell }
Similarly to the inductive case, we can implement coinductive types and
catamorphisms as follows:
\begin{minted}{haskell}
newtype CoFix f = OutOp {out :: f (Fix f )}
\end{minted}
As mentioned above, this type is similar to the data type
\mintinline{haskell}{Fix} we already defined. However, this time we are going to
use this assuming that it is the coinductive fixed-point by defining an
anamorphism on it:
\begin{minted}{haskell}
ana :: Functor b => (a -> b a) -> a -> CoFix b
ana coalg = OutOp . fmap (ana coalg) . coalg
\end{minted}
For example, the stream of natural numbers can now be defined using an anamorphism
\begin{minted}{haskell}
nat :: CoFix (ListF Int)
nat = ana (\x -> Cons x (x+1)) 0
\end{minted}
where the argument to the anamorphism is the
\mintinline{haskell}{ListF Int}-coalgebra.

@@

Let $(A,a)$ be an $F$-algebra and $(C,c)$ be an $F$-coalgebra. An arrow
$C \to A$ is an \emph{hylomorphism}, written $h : (C,c) \rightarrowtail (A,a)$ if
it satisfies
\begin{equation}
  \label{eq:hylo}
  h = a \circ F(h) \circ c
\end{equation}
As we stated earlier, a solution to this equation does not exist for an
arbitrary algebra and coalgebra pair and, in fact, this definition cannot be
accepted by Coq.

@@

\subsubsection{Haskell Implementation of Hylomorphisms}
% which is very useful in list processing. As a fold, The frequently used function
% map is also a fold:
% \begin{minted}{haskell}
% map :: (a -> b) -> [a] -> [b]
% map f = foldr (\x xs -> (f x):xs) []
% \end{minted}
We implement the hylomorphism tightly following the theory explained above:
\begin{minted}{haskell}
hylo :: Functor f => (c -> f c) -> (f a -> a) -> c -> a
hylo c a = a . fmap (hylo c a) . c
\end{minted}

A classic example of a divide-and-conquer algorithm that can be written as an
hylomorphism is the quicksort program below:
\begin{minted}{haskell}
partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r >= x])

quicksort :: [Int] -> [Int]
quicksort [] = []
quicksort (x:xs) = let (xs, x, ys) = partition x xs
                   in (quicksort xs) ++ [x] ++ (quicksort ys)
\end{minted}
The partition program takes an element \mintinline{haskell}{x} and a list and
returns a triple consisting of a list with all the elements in the list smaller
than \mintinline{haskell}{x}, the element itself and a list with all the
elements in the list greater or equal than  \mintinline{haskell}{x}. The
quicksort program first runs the partition program, which divides the list in
three parts, calls itself recursively on the sublists and then puts the results
together using the concatenation function.

To implement this as an hylo we define the functor representing trees where
leaves are just empty and each node contains a value:
\begin{minted}{haskell}
data TreeF a k = Leaf | Branch k a k deriving Functor
\end{minted}
Now the partition program can be written as a  \mintinline{haskell}{TreeF}-coalgebra
\begin{minted}{haskell}
divide :: [Int] -> TreeF Int [Int]
divide [] = Leaf
divide (p:xs) = Branch [l | l <- xs, l < p] p [r | r <- xs, r > p]
\end{minted}
and the concatenation function can be written as  \mintinline{haskell}{TreeF}-algebra
\begin{minted}{haskell}
conquer :: TreeF Int [Int] -> [Int]
conquer Leaf = []
conquer (Branch ls x rs) = ls ++ [x] ++ rs
\end{minted}
Finally, quicksort can be rewritten as an hylomorphism
\begin{minted}{haskell}
quicksort = hylo divide conquer
\end{minted}
As an important remark we have used the in-built list type for keeping the code
as tidy as possible, but it would have been possible to use the inductive
\mintinline{haskell}{List} type we defined above and define both
\mintinline{haskell}{divide} and \mintinline{haskell}{conquer} via
catamorphisms.

Finally, it is important to note that both  \mintinline{haskell}{divide} and
\mintinline{haskell}{conquer} are not initial algebras or terminal coalgebras on
\mintinline{haskell}{ListF}. These are rather (co)algebras on
\mintinline{haskell}{TreeF}.  Hence the terminating argument relies solely on
the fact that \mintinline{haskell}{divide} is a recursive coalgebra thus there
is an unique solution to \mintinline{haskell}{hylo divide conquer}.

\subsubsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive $F$-coalgebra and the
algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
from the uniqueness property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

@@

In this section, we focus on mechanising \emph{container functors} in a way that
is suitable for code extraction. Recall from Section~\ref{sec:rec-schemes} that
we need to be able to abstract away from particular shape of the data, and we
can do this using \emph{functors} that have suitable fixed-point properties;
i.e.\ those which have a initial algebras.
%
A common approach to construct such functors is to use
\emph{containers}~\cite{AbbottAG05}. However, reasoning about container equality
will require us to consider both functional extensionality and heterogeneous
equality. We avoid these axioms by introducing a custom equivalence relation on
types (Section~\ref{sec:types-morphisms}).

\subsection{Functors and Containers}
\label{sec:functors}

A straightforward mechanisation of functors as defined in
Section~\ref{sec:rec-schemes} does not work in Coq due to the strict positivity
condition. In fact, if we simply require that functors are functions
\coq{F : Type -> Type}
satisfying the necessary identity and composition properties, then we will not
be able to construct least/greatest fixed points of this type, and we will not
be able to mechanise recursion schemes that follow this structure:
\begin{coqcode}
(* REJECTED due to negative occurrence of (LFix F) *)
Inductive LFix (F : Type -> Type) := LFix_in (lfix_out : F (LFix F)).
\end{coqcode}

\noindent
One common workaround for this is the use of Mendler-style
recursion~\cite{Mendler91}. Roughly, the idea is to ``pack'' an abstract type
together with the functor that captures the structure of the recursion. From
this abstract type, we will be able to ``extract'' the contents of the functor.
This is, for example, the solution used by the PaCo library~\cite{PaCo}, and a
simplified version of it is the following:
\begin{coqcode}
Inductive MendlerLFix (F : Type -> Type) :=
  In Y (next : Y -> MendlerLFix F) (in_op : F Y).
\end{coqcode}

\noindent
While this definition solves the strict positivity problem, it does not solve
the problems of: (1) reasoning about data/program equalities; and (2) extraction
to idiomatic OCaml code. To reason about program equalities, we would need to
reason about equalities of functions with \emph{some abstract} \coq{Y} as their
domain.  Furthermore, the extraction of \coq{F Y} will lead to unsafe casts in
the extracted OCaml code.

% \paragraph{Containers}
Another common solution is to the strict positivity
problem is the use \emph{containers}~\cite{AbbottAG05}.  A container
$S \triangleright P$ is defined by a type of \emph{shapes} $S : \Type$ and a
\emph{family of position types} indexed by shapes $P : S \to \Type$.  An
\emph{extension} of this container is a functor
$\llbracket S \triangleright P \rrbracket$, whose action on objects is given by
$\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
and on morphisms is given by post-composition, keeping the shape the same
$\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.

To explain the role of shapes and positions in containers, consider the functor
$F\; X = 1 + A \times X$.  An equivalent container, i.e.\ a container with an
extension that is isomorphic to $F$, requires a shape that can distinguish two
cases, indicating that there are two constructors to build an object of type
$F\;X$. For example, we could define $S_{F} = 1 + A$. We need to consider two
cases for the family of positions:  $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because
there are no occurrences of $X$ in the left case; and
$P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence of $X$ in the
right hand side. It is easy to see that
$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.  In general, every
polynomial functor has an equivalent container, and we have mechanised this
result.
%
%Generally, any functor that
%is isomorphic to a container extension is called a \emph{container functor}.
%Here we will call \emph{container functor} the functors defined by a specific
%container extension.

\subsection{Extractable Containers}
A straightforward encoding of containers in Coq would use dependent types to
represent families of positions. However, Ocaml's type system is not equipped to
handle these, which will lead to extracting OCaml code with unsafe casts.

Consider instead representing the positions of a container using a decidable
validity predicate which assigns shapes to positions. In Coq, we use a boolean
function and a coercion from \coq{bool} to \coq{Prop} to represent the decidable
predicate \coq{valid}, similarly to SSReflect~\cite{GonthierL09}. Given a
\coq{Shape : Type}, and a type of \emph{all possible positions} \coq{APos : Type},
the family of positions for a given shape can be defined as a dependent pair:
\begin{coqcode}
Definition Pos (s : Shape) = { p : APos | valid s p }.
\end{coqcode}
Coq's code extraction will now be able to erase the validity predicate, and
generate OCaml code that is free of unsafe casts. The OCaml code extracted for
\coq{Pos} will now be exactly the code extracted for \coq{APos}.  The
decidability of the validity predicate is crucial for our purposes of  remaining
axiom-free. To illustrate this, suppose that we need to show the equality of two
container extensions. We will need to show that, for the same positions, they
will produce the same result. In Coq, the goal would look as follows:
\begin{coqcode}
 k : Pos s -> X
 P1, P2 : valid s p
 ---------------------------
 k (existT _ p P1) = k (existT _ p P2)
\end{coqcode}
If \coq{valid} was a regular proposition in \coq{Prop}, it would not possible
to prove the equality of \coq{P1} and \coq{P2}. However, by using a decidable
predicate, if we know that \coq{P1} and \coq{P2} are of type 
\coq{valid s p = true}, then we can prove without any axioms that 
\coq{P1 = P2 = eq_refl}.

Suppose that \coq{F} is a container represented in Coq. One can think of these
\coq{F} as pairs containing the shapes and families of positions (which are, in
turn, defined in terms of decidable validity predicates).
An extension of a container \coq{F}, \coq{App F} in Coq, is then defined as
follows:
\begin{coqcode}
Record App F (X : Type)
   := { shape : Shape F; cont : {p | valid shape p} -> X }.
\end{coqcode}

This approach solves the problem of the unsafe casts, since the code extracted
for \coq{{p | valid shape p}} will be an OCaml singleton type defined as the
OCaml equivalent to \coq{Pos}.

\subsubsection{Equality of Container Extensions}
Reasoning about the equality of container extensions is not entirely solved by
using decidable validity predicates to define the families of positions. In
general, we want to equate container extensions that have the same shape, and
that, for equal shape and position, they return the same element. To avoid the
use of the functional extensionality axiom, we capture this relation with the
following inductive proposition:
\begin{coqcode}
Inductive AppR (x y : App F X) : Prop :=
| AppR_ext (shape x = shape y)
           (forall p1 p2, projT1 e1 = projT1 e2 -> cont p1 = cont p2).
\end{coqcode}
Note that we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is trivially
\emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App F A -> B} (or \coq{B -> App F A}).
%
Since these types now come with their own equivalence, any function that
manipulates them needs to be \emph{respectful}; i.e.\ given
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{coqcode}
forall (x y : X), R x y -> R' (f x) (f y)
\end{coqcode}

>>>>

This paper presents the first Coq formalisation of a recursion scheme, called
the \emph{hylomorphism}, along with its algebraic laws allowing for the
mechanisation of all recognised (terminating) recursive algorithms. The key
contribution of this paper is that this formalisation is fully axiom-free
allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
framework by formalising a series of algorithms based on different recursive
paradigms such as divide-and conquer, dynamic programming, and mutual recursion
and demonstrate that the extracted OCaml code for the programs formalised in our
framework is efficient, resembles code that a human programmer would write, and
contains no occurrences of \ocaml{Obj.magic}.  Furthermore, we show the use of
the machine-checked proofs of the laws of hylomorphisms to do program
optimisations.
\end{abstract}

@@

There is not much tool support to use program calculation techniques. In fact,
uses of program calculation are done by  pen-and-paper proofs, and then
translating the result to specific instances of recursion schemes, implemented
in a programming language (generally Haskell).  There are some examples of
implementations of program calculation techniques, but these implementations are
scarce, not up to date, and not verified in a proof assistant. For example,
Cunha et al.~\cite{DrHylo} automated program calculation techniques by a custom
Haskell implementation.

Several authors build tools for applying program calculation techniques by
\emph{mechanising} them as part of a proof assistant. However, most of the
existing work only mechanises a narrow subset of the know recursion schemes, or
do not prove their algebraic
laws~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.  Indeed,
most of the mechanisations of generic recursion schemes cover \emph{maps} and
\emph{folds}, and few authors focus implement \emph{unfolds}. Omitting
\emph{unfolds} severely limits the expressivity of the resulting mechanisations,
and prevent them from being able to mechanise the most general recursion scheme:
a divide-and-conquer algorithm which goes under the name of
\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}. This generality has been proven by
Hinze et al.~\cite{HinzeWG15}, by showing that \emph{all known  recursion
schemes are instances of hylomorphisms}.  To date, no work has mechanised
hylomorphisms in the Coq proof assistant, together with their algebraic laws.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code, and it is not well-suited for
program calculation. This is unfortunate since code extration is what allows the
execution of code that has been verified in
Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
al's approach, extraction (1) does not produce code with the same recursive
structure  as existing manual implementations; and (2) leads to unsafe casts
like \ocaml{Obj.magic} in the generated code. This latter is also problematic in
that, for higher-order programs, simple interoperations can lead to incorrect
behaviour or even segfaults~\cite{forster:hal-04329663} and, moreover, it
invalidates the fast-and-loose principle~\cite{DanielssonHJG06}.

The contributions of this paper are as follows.  This work presents a Coq
formalisation of \emph{hylomorphisms} that (1) is \emph{fully axiom-free}; (2)
allows the extraction of idiomatic OCaml code; and (3) can use regular Coq
equalities to do program calculation, derive correct implementations, and apply
optimisations.  The full mechanisation is open source.
%can be found on
%Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
%  ANONYMISE THIS}.
While programmers still need to reason about the termination of their programs,
the use \emph{program calculation} can ease this task by allowing the use of
fusion on programs that have already been proven to terminate.

@@

\section{Recursion Schemes}\label{sec:rec-schemes}
The structure of data is very similar to the structure of an algorithm which
processes that data. This relationship manifests in the form of structured
recursion schemes which are widely used in functional languages such Haskell.
Canonical examples are folds (catamorphisms), which consume data, and unfolds
(anamorphisms), which produce it. While some implementations of recursion
schemes like $\operator{foldr}$/$\operator{unfoldr}$ in Haskell are specific to
a particular data structure, in this case $\operator{List}$s, we can generalise
further these ideas to account for generic (co)inductive data types and generic
algorithms operating on them.

@@

In $\Set$, we can define the set of lists
\[
  \List{A} \iso 1 + A \times \List{A}
\]
as the set inductively generated by the constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$.  The terminal object in $\Set$ is the
\emph{unit} type, that only has one inhabitant, $* : 1$.  The $\List{-}$ type is
a functor from  $\Set \to \Set$. When a functor maps a category to itself, it is
called an \textbf{endofunctor}. The action of $\List{-}$ is to take any set $A$
to the set $\List{A}$, and its functorial action
$\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(\nil(*)) = \nil(*)$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument.
However, there is a more general way of proving these type of definitions are
well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
the greatest set generated by the constructor
$\cons : A \times \Str{A} \to \Str{A}$.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
where $X$ is an object of the category called the \emph{carrier} of the algebra
and $a_{X}$ is an arrow of type $FX \to X$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
\emph{$F$-algebra homomorphisms}  $f : X \to Y$ in $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ a_{X} = a_{Y} \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
the initial $F$-algebra is precisely the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ with
$\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
isomorphism $F\mu F \iso \mu F$.  By the property of initial $F$-algebras, for
any other $F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra
homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (X, a_{X})$ and
pronounced ``\emph{catamorphism}'' satisfying:
\begin{equation}
  \label{eq:cata:unique}
  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InOp
\end{equation}
In words, any $F$-algebra homomorphism from the initial $F$-algebra is a
catamorphism. As a result of the uniqueness property we can derive the fusion
law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
\begin{equation}
  \label{eq:cata:fusion}
  f \circ \cata{a_{X}} = \cata{a_{Y}}
\end{equation}
which means that the composition of a program $f$ with a catamorphism recursing
once over the data structure is the same as performing that recursion once using
the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
optimisation as we shall see.

@@

@@

@@

Let $(A,a)$ be an $F$-algebra and $(C,c)$ be an $F$-coalgebra. An arrow
$C \to A$ is an \emph{hylomorphism}, written $h : (C,c) \rightarrowtail (A,a)$ if
it satisfies
\begin{equation}
  \label{eq:hylo}
  h = a \circ F(h) \circ c
\end{equation}
A solution to this equation does not exist for an arbitrary algebra and
coalgebra pair and, in fact, a definition like this cannot be accepted by Coq.

@@

\paragraph{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way. A recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible types it can produce from the seed are
the finite ones. We call this special kind of recursion scheme \emph{recursive
anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive $F$-coalgebra and the
algebra $\InIso$ for the inductive data type $\mu F$. This fact falls out
from the uniqueness property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

@@

A straightforward mechanisation of functors as defined in
Section~\ref{sec:rec-schemes} does not work in Coq due to the strict positivity
condition. In fact, if we simply require that functors are functions
\coq{F : Type -> Type}
satisfying the necessary identity and composition properties, then
we will not be able to construct least/greatest fixed points of this type.  A
common solution to the strict positivity problem is the use
\emph{containers}~\cite{AbbottAG05}.  A container $S \triangleright P$ is
defined by a type of \emph{shapes} $S : \Type$ and a \emph{family of position
types} indexed by shapes $P : S \to \Type$.  An \emph{extension} of this
container is a functor $\llbracket S \triangleright P \rrbracket$, whose action
on objects is given by
$\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
and on morphisms is given by post-composition, keeping the shape the same
$\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.
Intuitively, the type of shapes captures the different ``cases'' of an algebraic
datatype, and the positions of a shape are used to identify the elements inside
a container of the given shape.

Any polynomial functor can be represented as a container (see accompanying code
for a mechanisation).  For example, a container that is equivalent to
$F\; X = 1 + A \times X$ requires a shape that can distinguish two cases,
indicating that there are two constructors to build an object of type $F\;X$,
e.g. $S_{F} = 1 + A$. We need to consider two cases for the family of positions:
$P_{F}(\mathsf{inj}_{1}\;*) = 0$, because there are no occurrences of $X$ in the
left case; and $P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence
of $X$ in the right hand side. It is easy to see that
$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.

The straightforward mechanisation of containers and container extensions is
still not enough for our goals of remaining axiom-free, extracting to idiomatic
OCaml code, and avoiding \ocaml{Obj.magic} in the extracted code. The main
problem is that a container extension requires a function from a family of
positions to a type: $\forall s : S.\; P\;s \to X$. On the Coq side, dealing
with the equality of objects that contain (dependent) function types would force
us to add the functional extensionality axiom, and axiom K (or equivalent) to
deal with the heterogeneous equality of position.  For extraction, OCaml's type
system is not powerful enough to represent such dependencies in the general
case, which will lead to \ocaml{Obj.magic}.

To avoid the functional extensionality axiom, we require that all the types used
in our definitions are \emph{setoids}, where we will use \coq{x =e y} to
represent setoid equality. All morphisms need to be \emph{proper} morphisms with respect
to the function \emph{respectfulness} relation. In our development, we use
\begin{coqcode}
Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
\end{coqcode}
We add an automatic coercion from \coq{A ~> B} to \coq{A -> B}, provide tactics
to automatically discharge proofs that \coq{forall x y, x =e y -> f x =e f y}
for the simplest cases of \coq{f}, and we provide a library with
definitions for basic building blocks: sum and product types, exponentials,
composition, identity,
\ldots

To avoid axiom K, and extract containers to idiomatic OCaml code, we
define containters in terms of shapes, \emph{base} positions (\coq{APos : Set}),
and \emph{decidable validity predicates} that
specify when a base position is valid in a shape. Since validity predicates are in
\coq{Prop}, they will be erased during extraction, and their decidability
will make unnecessary the use of axiom K.
\begin{coqcode}
Record App F (X : Type) := {shape : Shape F; cont : {p : APos | valid s p = true} -> X}.
\end{coqcode}

Since we require every type to be a setoid, we need to define a setoid equality
(\coq{=e}) for container functors. Given two \coq{x y : App F X}, we define
\coq{x =e y} as the setoid equality of the shapes, and the extensional equality of
the contents, given equal shapes. I.e.\ \coq{x =e y} iff
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> cont x p =e cont y p')
\end{coqcode}
